# Data-generating process functions

In the last chapter, you learned how to write functions. In this chapter, you’ll use that skill to write a specific type of function: **a data-generating process (DGP) function**. This is needed for one of the 5 key components of a Monte Carlo simulation study.

::: {.callout-note collapse="true"}
## Can you recall the 5 key components of Monte Carlo simulation?
1. Design an experiment  
2. Generate data  
3. Analyse data  
4. Do this many times  
5. Summarise results
:::

## What is a data-generating process?

A **data-generating process (DGP)** is the theoretical or mathematical mechanism that causally produces observed data. It is the "ground truth". 

- In a real-world study, the DPG is unknown and not directly knowable, but may be estimated from empirical data. For example, the heights of humans in the population appear to follow a normal distribution with a given population mean and standard deviation.
- In a simulation, we have access to the ground truth because we define the data-generating process.

A data-generating process is defined by a **probability density function** and its specific **parameters**. Put simply, these define what is "true" in the world that we are simulating. In psychological research, we are most familiar with the normal (or Gaussian) probability density function, but there are many others.

For example, let's imagine we want to simulate a dataset of the heights of adult women versus men. We can take data from the US Center for Disease Control's NHANES dataset, in this case for the years 2008-2012, which contains data for about 7000 representative US adults.

Let's plot the data:

```{r}
#| echo: false

# dependencies
library(dplyr)
library(tidyr)
library(tibble)
library(roundwork)

# NHANES dataset from the US Center for Disease Control, US adults 2008-2012
library(NHANES)
data(NHANES)

# wrangle data
data_height <- NHANES |>
  janitor::clean_names() %>%
  filter(age >= 18) %>%
  filter(!is.na(height)) %>%
  select(gender, height)

# summarize
data_height_summary <- data_height %>%
  group_by(gender) %>%
  summarize(
    n = n(),
    mean_height = mean(height),
    sd_height = sd(height),
    mean_minus_sd = mean(height) - sd(height),
    mean_plus_sd = mean(height) + sd(height)
  )

ggplot(data_height, aes(height, fill = gender)) +
  geom_density(adjust = 2, alpha = 0.4) +
  theme_linedraw() +
  labs(title = "Distribution of heights in US adults (2008-2012) by gender",
       x = "Height",
       y = "Density") +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1")
```

We can calculate the means and standard deviations:

```{r}
#| echo: false

data_height_summary |>
  select(gender, mean_height, sd_height) %>%
  mutate_if(is.numeric, round_up, digits = 0)
```

This is a very common step in psychological research - so common that it's easy to overlook something subtle: means and standard deviations are the parameters that define a normal distribution (i.e., and not other distributions). 

By calculating means and standard deviations, we are implicitly employing a normal probability density function, and endorsing the idea that these are a meaningful summary of the data and, more importantly, a description of the data generating function.

Let's check whether that is accurate. For this simple example, we can do this by checking whether a normal distribution with these parameters fits the observed data.

```{r}
#| echo: false
ggplot(data_height, aes(x = height, fill = gender, color = gender)) +
  # Empirical Density
  geom_density(adjust = 2, alpha = 0.4, color = NA) +
  # Normal Distribution Overlay
  stat_function(fun = dnorm, 
                args = list(mean = data_height_summary$mean_height[data_height_summary$gender == "female"], 
                            sd = data_height_summary$sd_height[data_height_summary$gender == "female"]),
                color = "#E41A1C", linetype = "dashed", linewidth = 1) + 
  stat_function(fun = dnorm, 
                args = list(mean = data_height_summary$mean_height[data_height_summary$gender == "male"], 
                            sd = data_height_summary$sd_height[data_height_summary$gender == "male"]),
                color = "#377EB8", linetype = "dashed", linewidth = 1) +
  theme_linedraw() +
  labs(title = "Distribution of heights in US adults (2008-2012) by gender",
       subtitle = "Dashed lines represent Normal distributions using the empirical Means and SDs",
       x = "Height",
       y = "Density") +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1")
```

They do. We can therefore say that this probability density function and these parameters represent a good estimate of the DGP. 

These values for the means and standard deviations represent the (assumed) **population** parameters - i.e., the ground truth. If we want to simulate actual datasets - small samples drawn from this population - how would we do this? We would need to write a function that draws such random samples from this population, i.e., following this DGP. This is analogous to recruiting samples from a population in a real study. 

Remember that functions have three components:

1. They take inputs
2. They do stuff to the inputs
3. They return outputs

A DGP function's properties are, then, more specifically:

1. Take **parameters** as inputs
2. Draw samples from a DGP based on a given probability density function and the input parameters
3. Return a simulated dataset

## Building our first data-generating function: simulating two groups drawn from a normal distribution

A simple example that is useful to many simulations of psychological data - and the example that we're going to use frequently in the next several chapters - is the simulation of two groups drawn from normal distributions with known means and standard deviations. In the previous example, these referred to heights for men and women, but in many of the following examples they will instead represent the intervention vs. control group of a randomised controlled trial (e.g., psychotherapy for depression). 

Let's think about our inputs first. What types of things might vary in such a design? The obvious choice is the means in the control and intervention groups. We could also vary the standard deviation (SD) associated with the means. And we could also vary the number of participants we have in each condition. This means we have as inputs:

- `n_per_condition`: sample size per group
- `mean_control`: mean in the control group
- `mean_intervention`: mean in the intervention group
- `sd`: within-group standard deviation

We will use R's `rnorm()` function to generate the data for each group. `rnorm()` takes three arguments: `n` (the sample size), `mean` (the mean value), and `sd` (the standard deviation). `rnorm()` will then give us `n` random draws from a normal distribution with a mean of `mean` and an SD of `sd`.

Let's remember the steps to write a function that we learned in the previous chapter: write the 'do stuff' (as pseudocode first if needed) and extract hard-coded values as variables, then wrap it in a function.

### Step 1: write the 'do stuff'

Start by getting the data generation working with hard-coded values. Pseudocode for this could look like:

```default
1. Sample data from a normal distribution using `rnorm()`
2. Sample data for two different groups using defined N, M, SD.
3. Label the groups intervention and control
4. Return a single tibble with the combined data
```

Start as simply as possible: Get `rnorm()` working by itself.

```{r}
#| include: false
# set the seed - invisibly for the moment - seeds will be explained later
set.seed(42)
```

```{r}
rnorm(n = 20, 
      mean = 28, 
      sd = 8)
```

Return the results as a tibble rather than a vector. We will return to this point about the importance of tidy inputs and outputs later.

```{r}
tibble(score = rnorm(n = 20, 
                     mean = 28, 
                     sd = 8))
```

Replace the hard-coded values with variables.

```{r}
n_per_condition <- 20
mean_intervention <- 28
sd <- 8

tibble(score = rnorm(n = n_per_condition, 
                     mean = mean_intervention, 
                     sd = sd))
```

Create two different tibbles, one for each group. The appropriate population mean is used in each.

```{r}
n_per_condition <- 20
mean_intervention <- 28
mean_control <- 32
sd <- 8

data_intervention <- 
  tibble(score = rnorm(n = n_per_condition, 
                       mean = mean_intervention, 
                       sd = sd))

data_control <- 
  tibble(score = rnorm(n = n_per_condition, 
                       mean = mean_control, 
                       sd = sd))
```

Add a condition variable to each, so you know which condition each score belongs to.

```{r}
n_per_condition <- 20
mean_intervention <- 28
mean_control <- 32
sd <- 8

data_intervention <- 
  tibble(condition = "intervention",
         score = rnorm(n = n_per_condition, 
                       mean = mean_intervention, 
                       sd = sd))

data_control <- 
  tibble(condition = "control",
         score = rnorm(n = n_per_condition, 
                       mean = mean_control, 
                       sd = sd))
```

Bind the two tibbles together to make a single dataset.

```{r}
n_per_condition <- 20
mean_intervention <- 28
mean_control <- 32
sd <- 8

data_intervention <- 
  tibble(condition = "intervention",
         score = rnorm(n = n_per_condition, 
                       mean = mean_intervention, 
                       sd = sd))

data_control <- 
  tibble(condition = "control",
         score = rnorm(n = n_per_condition, 
                       mean = mean_control, 
                       sd = sd))

data_combined <- bind_rows(data_intervention, 
                          data_control)
```

If you `View(data_combined)`, you can see this looks like our desired output. You have written the 'do stuff'!

Or have we? Let's do some sanity checks before we go further. Once it's wrapped as a function, the inner workings of the intermediate steps become somewhat invisible due to varible scoping (see previous chapter). It's easier to verify it first.

Let's increase `n_per_condition` to be large, so that the sample mean and standard deviations are closer to the population values.

```{r}
n_per_condition <- 10000
mean_intervention <- 28
mean_control <- 32
sd <- 8

data_intervention <- 
  tibble(condition = "intervention",
         score = rnorm(n = n_per_condition, 
                       mean = mean_intervention, 
                       sd = sd))

data_control <- 
  tibble(condition = "control",
         score = rnorm(n = n_per_condition, 
                       mean = mean_control, 
                       sd = sd))

data_output <- bind_rows(data_intervention, 
                         data_control)
```

Now, let's check that our single simulated dataset's sample mean and sample standard deviation are a close match for the DGP's population mean and population standard deviation.

If all is in order, `n_per_condition` should be 10000 for each condition; the mean of `score` should be roughly 32 for the control condition and roughly 28 for the intervention condition; and SDs should be roughly 8 for both conditions. 

```{r}
data_combined |>
  group_by(condition) |>
  summarise(n_per_condition = n(),
            mean = mean(score),
            sd = sd(score),
            .groups = "drop") |>
  mutate_if(is.numeric, round_up, digits = 0)
```

They match! This might seem trivial at first, but it is a) an important sanity check when writing functions, and b) is the basis of **parameter recovery** which we will return to later in the book.

### Step 2: wrap it into a DGP function

Now let's turn our working code into a reusable DGP function. 

Move the 'do stuff' code we just wrote into the function skeleton.

```default
function_name <- function(inputs){
  # 'do stuff' goes here
  return(output)
}
```

```{r}
generate_data <- function(n_per_condition,
                          mean_control,
                          mean_intervention,
                          sd) {
  
  data_control <- tibble(
    condition = "control",
    score = rnorm(n = n_per_condition, 
                  mean = mean_control, 
                  sd = sd)
  )

  data_intervention <- tibble(
    condition = "intervention",
    score = rnorm(n = n_per_condition, 
                  mean = mean_intervention, 
                  sd = sd)
  )

  data_output <- bind_rows(data_control, data_intervention) 

  return(data_output)
}
```

Now let's call the `generate_data` function and see if it works.

```{r}
generate_data(
  n_per_condition = 10000,
  mean_control = 28,
  mean_intervention = 32,
  sd = 8
)
```

If it didn't, how can you bug-fix it? Go back to the last chapter for common errors.

Now sanity check it again. As we discussed in the last chapter, there is a difference between code that runs and code that functions as intended. 

A simple sanity check here is to change the inputs and check that the outputs change appropriately.

```{r}
# population parameters
results1 <- generate_data(
  n_per_condition = 5000,
  mean_control = 76,
  mean_intervention = 12,
  sd = 33
)

# attempt to recover those parameters in a large sample
results1 |>
  group_by(condition) |>
  summarise(n_per_condition = n(),
            mean = mean(score),
            sd = sd(score),
            .groups = "drop") |>
  mutate_if(is.numeric, round_up, digits = 0)
```

## Why return tidy data?

In the next chapter, you’ll write **analysis functions** that take a dataset and return a result (a *p*-value, an estimated effect, a confidence interval, etc). When we write these analysis functions, we will have certain expectations about the output of our data-generating functions (e.g., that they will conform to principles of tidy data).

In order to ensure that every data-generating function returns a tidy dataset with predictable column names, it is good practice to make sure that every DGP function returns a tibble as output. Tibbles are like dataframes, but with more consistent behavior that is useful for analysis. For example, if we use `select()` and select a single column from a dataframe, this is automatically treated by R as a vector, which might break some of our analysis functions down the line (which expect tibbles). `select()`ing one column from a tibble, by contrast, still returns a tibble object.

Importantly, the outputs of our DGP functions should usually consist of the following information:

- one row per observation
- relevant grouping variables
- one column per variable you'll analyse

## Common data-generating functions

In practice, many of the data-generating processes that we want to simulate can be modelled using some useful functions in R. You have already seen one example of this used above: namely, `rnorm()`. In this section, we'll go through a small number of additional data-generating functions that you will likely find useful in your future simulations.

### Simulating binary outcomes with `rbinom()`

`rbinom()` is used to model binomial distributions. This might sound a little offputting, but it's actually very simple: binomial distributions refer to cases where there are only two possible answers. Some examples of such situations are heads/tails in coin-flips, correct/incorrect in testing, dropped out/finished in an intervention study, and many others. 

Let's look at how to use `rbinom()`. It takes three arguments: `n`, `size`, and `prob`. `n` refers to the number of samples (e.g., how many total rounds of coin-flips are done); `size` refers to the number of observations per sample (e.g., 10 coin-flips are done for every round of coin-flips collected); and `prob` refers to the "probability of success" (e.g., in coin-flips both outcomes are equally likely, so the probability of "success" (defined either as getting heads or getting tails) is 0.5).

Let's look at a simple example: we want to simulate 10 rounds of flipping a single coin, assuming a probability of success of 0.5 for that coin. 

```{r}

set.seed(42)
rbinom(n = 10, size = 1, prob = 0.5)

```

Let's expand this into our two-group example from before. Previously, we used `rnorm()` to simulate values for a continuous outcome; now we can simulate binary values (e.g., getting a test question correct, scored as 1, or incorrect, scored as 0) with differences in the probability of correct responses between the two groups. This looks like:

```{r}

generate_data_two_group_binary <- function(n_per_condition,
                                           prob_control,
                                           prob_intervention) {

  data_control <- tibble(
    id = 1:n_per_condition,
    condition = "control",
    score = rbinom(n_per_condition, size = 1, prob = prob_control)
  )

  data_intervention <- tibble(
    id = (n_per_condition + 1):(2 * n_per_condition),
    condition = "intervention",
    score = rbinom(n_per_condition, size = 1, prob = prob_intervention)
  )

  dat <- bind_rows(data_control, data_intervention) |>
    as_tibble()

  return(dat)
}

```

We can check the observed proportions to make sure our functions is working correctly. Specifically, if we simulate with `n_per_condition` = 30, `prob_control` = 0.25, and `prob_intervention` = 0.75, then we should see these proportions in the estimated mean scores of each group:

```{r}

set.seed(42)

data_bin <- generate_data_two_group_binary(
  n_per_condition = 30,
  prob_control = 0.25,
  prob_intervention = 0.75
)

data_bin |>
  group_by(condition) |>
  summarise(estimated_mean_score = mean(score), .groups = "drop")

```

Using `rbinom()` is crucial for simulating binary outcome DGPs. Later, your analysis function might be use a logistic regression to analyse these binary data. But note that, even though the specific DGP function has changed, the *structure* of the `generate_data_two_group_binary` function looks the same as `generate_data_two_group` from before: take inputs -> do stuff -> output a tibble.

### Correlated data: repeated measures

Psychology data often include repeated measurements on the same person. For example, we might measure a person's score on a scale before and after they receive an intervention. In principle, we could just simulate these observations using two different `rnorm()` calls, like this:

```{r}

generate_data_two_timepoints <- function(n_per_timepoint,
                                    mean_score_t1,
                                    mean_score_t2,
                                    sd) {

  dat <- tibble(
    id = 1:n_per_timepoint,
    score_t1 = rnorm(n_per_timepoint, mean = mean_score_t1, sd = sd),
    score_t2 = rnorm(n_per_timepoint, mean = mean_score_t2, sd = sd)
  )

  return(dat)
}


```

But let's run this function and look at the correlation between these observations:

```{r}

set.seed(42)

two_timepoint_data <- generate_data_two_timepoints(n_per_timepoint = 50,
                                                   mean_score_t1 = 0,
                                                   mean_score_t2 = 0.5,
                                                   sd = 1)

broom::tidy(
  cor.test(two_timepoint_data$score_t1,
           two_timepoint_data$score_t2)
)

```

The scores from timepoint 1 (T1) and timepoint 2 (T2) are uncorrelated. But in real life, this rarely happens: two observations from the same participant across time tend to be correlated, even when they are influenced by an intervention. 

So, how do we simulate *correlated* variables?

One simple way to do this is to draw from what is known as a **multivariate normal distribution**. This is a distribution that lets us specify both the means and SDs of variables, as well as the correlation between them. The function `mvrnorm()` from the {MASS} library does this very well.

Let's simulate a design where we have a pre-intervention score variable `mean-pre`, and a post-ntervention score variable `mean-post`, and a correlation between these two variables `cor_pre_post`. 

To use a multivariate normal distribution, we need to specify the *covariance matrix* between the variables. This is just a matrix that specifies how related each variable is to each other. Since we have two variables here, we need to create a 2x2 matrix. 

You can think of a covariance matrix as an unstandardised correlation matrix. The correlation between a variable and itself is always 1. The covariance between a variable and itself is always the square of its standard deviation, namely `sd^2`. 

When the SD is the same at both time points, the covariance between two variables is the correlation between them multiplied by `sd^2`, AKA `cor_pre_post * sd^2`. When we use the `mvrnorm()` function, we supply this covariance matrix to it as the parameter `Sigma`. 

Our simple multivariate DGP function is as follows:

```{r}

generate_data_pre_post <- function(n,
                                   mean_pre,
                                   mean_post,
                                   sd,
                                   cor_pre_post) {

  # construct the covariance matrix
  covariance_matrix <- matrix(
    c(sd^2, cor_pre_post * sd^2,
      cor_pre_post * sd^2, sd^2),
    nrow = 2,
    byrow = TRUE
  )

  # simulate from the multivariate normal distribution
  raw_dat <- MASS::mvrnorm(
    n = n,
    mu = c(mean_pre, mean_post),
    Sigma = covariance_matrix
  )


  dat <- as_tibble(raw_dat) |>
    setNames(c("score_pre", "score_post")) |>
    mutate(id = 1:n) |> # create an identifier
    relocate(id)

  return(dat)
}

```

And here's what the output of this looks like:

```{r}

set.seed(42)

data_pre_post <- generate_data_pre_post(
  n = 50,
  mean_pre = 0,
  mean_post = 0.5,
  sd = 1,
  cor_pre_post = 0.7
)

data_pre_post

```

Once again, we can sanity-check our code to make sure we see both means and between-variable correlations like we expect:

```{r}

data_pre_post |>
  summarise(mean_pre = mean(score_pre),
            mean_post = mean(score_post),
            cor_pre_post = cor(score_pre, score_post))

```

### Many variables at once: `faux::rnorm_multi()`

{MASS}'s `mvrnorm()` function is handy for generating multiple correlated variables. However, you may have noticed a couple of things that are not ideal about it. First, we need to manually construct the covariance matrix, which requires a little bit of fiddling and is generally a bit of a pain - especially if we want to simulate many variables, it can become quite a lot of work. Second, `mvrnorm()` does not give tidy output, meaning we have to do extra data cleaning and processing. 

Fortunately, there exists a much handier function for simulating multivariate normal data: `rnorm_multi()` from the {faux} R package. We can use `rnorm_multi()` to simulate many correlated variables at once. This is especially useful for tasks like generating many items of a scale which should be correlated.

If you don’t have it installed yet, install it once:

```{r}
#| eval: false

install.packages("faux")

```

`rnorm_multi()` can take a few different arguments. They are:

- `n` = the sample size per variable
- `vars` = the number of variables to be generated
- `mu` = the means of the variables
- `sd` = the SDs of the variables
- `r` = the correlation between each of the variables
- `varnames` = the desired names of the variables
- `empirical` = whether the correlations and means should exactly match the input values or not
- `as.matrix` = whether the output should be returned as a matrix (TRUE) or not (FALSE)

Obviously, we want our output as a tibble, not a matrix, so we will set `as.matrix` to FALSE. This is actually the default value, so we can leave it alone.

Let's use `rnorm_multi()` to generate 5 variables, each with 200 observations `n`, each with a mean `mu` of 0, a sd `sd` of 1, and a correlation `r` of 0.3. 

```{r}

set.seed(42)

data_simulated_items <- faux::rnorm_multi(
  n = 200,
  vars = 5,
  mu = 0,
  sd = 1,
  r = 0.3,
  varnames = c("variable1", "variable2", "variable3", "variable4", "variable5")
) |>
  as_tibble()

```

Let's look at the correlations between the simulated items:

```{r}

cor(data_simulated_items)

```

They look roughly right. But we can use the `empirical` argument to make them more precise. Let's try this out again with `empirical = TRUE`:

```{r}

set.seed(42)

data_simulated_items_empirical <- faux::rnorm_multi(
  n = 200,
  vars = 5,
  mu = 0,
  sd = 1,
  r = 0.3,
  varnames = c("variable1", "variable2", "variable3", "variable4", "variable5"),
  empirical = TRUE
) |>
  as_tibble()

```

Let's look at the correlations between the simulated items:

```{r}

cor(data_simulated_items_empirical)

```

Setting `empirical = TRUE` can make the simulation take longer to run, but it can also provide us with data that more precisely adheres to our specified parameter values. Although we usually do not want this in a simulation where we do this process many times.

A full DGP might simulate these items, and then compute a total score based on these individual items:

```{r}

generate_data_scale_score <- function(n, 
                                mean, 
                                sd, 
                                n_items,
                                r) {

  items <- faux::rnorm_multi(
    n = n,
    vars = n_items,
    mu = mean,
    sd = sd,
    r = r,
    varnames = paste0("variable", 1:n_items)
  ) |>
    as_tibble()

  dat <- tibble(
    id = 1:n,
    score = rowMeans(items)
  )

  return(dat)
}

```

Let's run this function where we assume 100 participants completed 10 items, each item having a mean of 0, SD of 1, and a correlation with the other items of 0.5:

```{r}

generate_data_scale_score(n = 100, 
                          mean = 0, 
                          sd = 1, 
                          n_items = 10,
                          r = 0.5)

```

What is returned, as we specified in our function above, is a single score for each participant, representing the mean of the 10 item scores for that participant. But we know that these scores are produced based on 10 items with a mean of 0, SD of 1, and correlation of 0.5 among them. 

### Generating variables with different correlations

Above, we looked at how we can use `rnorm_multi()` to generate multiple variables all with the same mean, SD, and correlation between them. But we can actually also use `rnorm_multi()` to generate data where variables have different means, SDs, and inter-variable correlations.

The trick is that the `r` argument does not have to be a single number. It can also be a **correlation matrix**: a square matrix with:

- 1s on the diagonal (each variable correlates perfectly with itself)
- your desired correlations in the off-diagonal cells

This is very similar to what we did previously with `mvrnorm()`, but with one useful difference: we can specify the correlation matrix directly, instead of the covarianece matrix (which is harder to understand since it is unstandardised).

This is easiest to see with a small example. Suppose we want three variables with these correlations:

- cor(x, y) = 0.6
- cor(x, z) = 0.2
- cor(y, z) = 0.4

We can write that as a correlation matrix:

```{r}

cor_matrix <- matrix(
  c(1, 0.6, 0.2,
    0.6, 1, 0.4,
    0.2, 0.4, 1),
  nrow = 3,
  byrow = TRUE
)

cor_matrix

```

Now we can pass that matrix into `rnorm_multi()`. Here we also set different means and SDs for each variable by passing vectors to `mu` and `sd`. We set `empirical = TRUE` to illustrate that the sample correlations match what we specified exactly.

```{r}

set.seed(42)

data_different_correlations <- faux::rnorm_multi(
  n = 200,
  vars = 3,
  mu = c(0, 0.5, -0.25),
  sd = c(1, 1, 1.5),
  r = cor_matrix,
  varnames = c("x", "y", "z"),
  empirical = TRUE
) |>
  as_tibble()

data_different_correlations |>
  summarise(
    mean_x = mean(x),
    sd_x = sd(x),
    mean_y = mean(y),
    sd_y = sd(y),
    mean_z = mean(z),
    sd_z = sd(z)
  )

cor(data_different_correlations)

```

This same idea can scale up. For example, we can build a “two-subscale” correlation structure for six items: items 1–3 correlate highly with each other (within-subscale), items 4–6 correlate highly with each other, and correlations across subscales are weaker. Think about this in the context of scales like personality questionnaires, which usually have 5 "subscales" where items correlate strongly with one another within subscales, and more weakly across subscales.

```{r}

n_items <- 6
r_within <- 0.6
r_between <- 0.2

cor_items <- matrix(r_between, nrow = n_items, ncol = n_items)
diag(cor_items) <- 1
cor_items[1:3, 1:3] <- r_within
cor_items[4:6, 4:6] <- r_within
diag(cor_items) <- 1

cor_items

```

```{r}

set.seed(42)

data_two_subscales <- faux::rnorm_multi(
  n = 200,
  vars = n_items,
  mu = 0,
  sd = 1,
  r = cor_items,
  varnames = paste0("item", 1:n_items),
  empirical = TRUE
) |>
  as_tibble()

cor(data_two_subscales)

```

::: {.callout-note collapse="true" title="If you get an error about the correlation matrix"}
Not every set of correlations is mathematically possible. If `rnorm_multi()` tells you the correlation matrix is “not positive definite”, it usually means your correlations are too extreme or inconsistent with each other. A simple fix is to reduce some of the correlations (move them closer to 0).
:::

## Pseudo-random number generators

Randomness is impossible to achieve. All "random" number generators are actually pseudo-random number generators (PRNGs). Computer scientists and mathematicians spend lots of time trying to increase the randomness of our random number generators, because any degree of predictability adds bias to any models built on them. Pseudo-random numbers are at the core of simulation studies, as they allow us to (pseudo) randomly sample simulated data from known population distributions.

### Sampling from uniform distributions

A uniform distributions is when every value is as likely as every other value, and are selected from a given range. E.g., "pick a number between 1 and 10" where the picker is just as likely to say "1" as any other number in that range. 

`runif()` is a random generation function (the "r" part) for a uniform distribution (the "unif" part). Not 'run if', which confuses some people.

So, this code, which generates a random number between 1 and 10, will generate 3s just as often as it generates 7s. You can re run this code yourself many times to see it generate different numbers between 1 and 10.

```{r}


runif(n = 1, 
      min = 1,
      max = 10) |>
  round_up(0)

```

### Random number generators are not truly random

You don't need to understand how PRNGs work, but you do need to know that the these "random" numbers can be predictably reproduced. The 'seed' value from which a PRNG starts can be set to control which random numbers are generated. 

```{r}

set.seed(43) # set the starting seed value for generating the random numbers

runif(n = 1, 
      min = 1,
      max = 10) |>
  round_up(0)


set.seed(43) # set it again to the same value starting seed value for generating the random numbers

runif(n = 1, 
      min = 1,
      max = 10) |>
  round_up(0)

```

Note that if you run the function a second time without resetting to a known seed, the second value will be different to the first one.  

```{r}

set.seed(43) # set the starting seed value for generating the random numbers

runif(n = 1, 
      min = 1,
      max = 10) |>
  round_up(0)

runif(n = 1, 
      min = 1,
      max = 10) |>
  round_up(0)

```

This is because the Nth value of any sequence from a given seed is knowable, whether its run once or in multiple runs.

```{r}

set.seed(43) # set the starting seed value for generating the random numbers

# generate both of the above numbers in one function call
runif(n = 2, # generate two numbers rather than one
      min = 1,
      max = 10) |>
  round_up(0)

```

### Randomness and reproducibility

You might have heard people talking about "seeds" before when using R. Seeds, basically, are used to make sure that random output can be reproducible. 

For example, look at the outputs of two calls to the `rnorm()` function below, which samples random values from a normal distribution with a mean of 0 and a standard deviation of 1:

```{r}

output1 <- rnorm(n = 5, mean = 0, sd = 1)
output2 <- rnorm(n = 5, mean = 0, sd = 1)

output1 
output2

```

You can see that the outputs of the two calls are different. But sometimes we might not want this: for example, when we want other people to be able to reproduce our simulation results exactly. If we use truly random numbers, then every time we re-run the simulation we'll get slightly different results. What we really want are *pseudo*-random numbers - numbers which are sampled randomly, but where the specific random sampling can be reconstructed and reproduced. This is exactly what seeds are for: if you set a seed to a specific value, then all calls to functions which use randomisation will reproduce their random samples in future runs. 

Put simply: if you start from the same seed, you get the same sequence of "random" values. Check out the example below, now using `set.seed()`:

```{r}

set.seed(123)

output1 <- rnorm(n = 5, mean = 0, sd = 1)
output2 <- rnorm(n = 5, mean = 0, sd = 1)

output1 
output2

```

In your own work, you’ll often set a seed once at the top of a script so you (and others) can reproduce your results.

::: {.callout-warning collapse="true" title="A common mistake: setting the seed inside your generator"}
Don’t put `set.seed()` *inside* a DGP function. If you do, every call will restart the random number generator in the same place, and you will keep generating the same dataset. This becomes a problem when we get to Step 4 (repeat many times), since you just end up repeating the exact same process every time.

```{r}

bad_generate <- function() {
  set.seed(42)
  rnorm(3)
}

bad_generate()
bad_generate()

```

```{r}

set.seed(42)
good_generate <- function() {
  rnorm(3)
}

good_generate()
good_generate()

```
:::


## A checklist for DGP functions

When you write your own `generate_data()` functions, aim for these defaults:

- Put all "design choices" in the arguments (sample sizes, effect sizes, standard deviations, etc.).
- Return **one** dataset per function call.
- Avoid side effects (don’t call `set.seed()` inside the function; don’t write files, don’t make plots).
- Use clear, consistent column names (which helps both for reading code and for running analysis functions later).

## Summary

- A **data-generating process** is a set of assumptions plus randomness introduced by the sampling.
- In simulation, parameters define what is "true" and random draws define what happens in one sample.
- A DGP function should take explicit inputs and return one tidy dataset.
- Useful building blocks include `rnorm()` (continuous), `rbinom()` (binary), and multivariate normal draws for correlated variables (`mvrnorm()`, `rnorm_multi()`).



